Towards understanding the central limit theorem by learning Python basics

### 2021-02-19 paper/article.tex
Add additional text, figures and program translations

>>>>

\subsection*{Sheet 5}

Fig.~\ref{5a.py} shows the first program in the fifth sheet. The students' assignment was
to find a text problem that is solved by the program.

\begin{figure}
\lstinputlisting[language=Python]{../progs/en_US/5a.py}
\caption{The first program in the fifth sheet (translated into English)}
\label{5a.py}
\end{figure}

To find a connection with binomial coefficients, the students had to explain why the result
equals to $\binom{5}{2}/2^5$. A technical question helped in improving the look of the output,
by suggesting the addition of this line between lines 2 and 3:
\begin{lstlisting}[language=Python,numbers=none]
Ω = {"".join(outcome) for outcome in Ω}
\end{lstlisting}

Then a second program was introduced. In fact, it was just a slight modification of the first program,
but, for simplicity, the whole code was announced (see Fig.~\ref{5b.py}).
\begin{figure}
\lstinputlisting[language=Python]{../progs/en_US/5b.py}
\caption{The second program in the fifth sheet (translated into English)}
\label{5a.py}
\end{figure}
Here the assignment was to give a random variable $X$ whose expected value is computed in the program.
Another assignment was to explain why the expected value $2.5$ is printed. Then, this result
had to be generalized by changing the code \texttt{repeat=5} to other values. Finally,
also the variance had to be computed by adding some more lines to the code.

The first program was well identified by all students---a fair coin was tossed 5 times,
and the probability of the appearance of exactly two heads was computed.
Applying the technical hint was unsuccessful for two students (for one the closing brace was missing,
for another one the issue was unclear).

For the second program, one student added the line \texttt{print(len($\Omega$))} to better understand the
code. (In fact, the function \texttt{len} was already introduced in the third sheet.) Instead
of mentioning symmetry, the same student argued by citing the formula $E=n\cdot p$ for the
binomial distribution where $n$ denotes the number of independent experiments and $p$ the probability
of success (here $0.5$). The same student computed the variance by multiplying $E$ with $0.5$---this
is correct for the given assignment, but here no programming was used, just theory.

The other students explained the result $2.5$ by mentioning symmetry, and the same argument was
used for the generalization. But no answer was found to compute the variance, only for one student,
and in that case it was incorrect (see Fig.\ref{52})---clearly, on line 10 something like
\begin{lstlisting}[language=Python,firstnumber=10]
  Varianz = Varianz + (w.count("K")-Erwartungswert)**2
\end{lstlisting}
should have been written. Actually, the notation \texttt{**2} could be something unfamiliar
for those who never did any programming before. Anyway, the result $0.078125$ should have been
suspicious. It is also remarkable that this student preferred the Latin alphabet to Greek characters.
\begin{figure}
\begin{center}\includegraphics[width=1.0\textwidth]{52}
\caption{A student's incorrect program code to compute variance in sheet 5}
\label{52}
\end{center}
\end{figure}

\subsection*{Sheet 6}

We identified the assignment to compute the variance in the fifth sheet somewhat unsuccessful.
So we continued by sharing two correct solutions to compute the variance (see Fig.~\ref{6a.py}
and \ref{6b.py}).

\begin{figure}
\lstinputlisting[language=Python]{../progs/en_US/6a.py}
\caption{A correct solution to compute the variance (translated into English)}
\label{6a.py}
\end{figure}

\begin{figure}
\lstinputlisting[language=Python]{../progs/en_US/6b.py}
\caption{Another correct solution to compute the variance (translated into English)}
\label{6b.py}
\end{figure}

The students had to compare these two program codes, check if their outputs are the same
(even for 6 or 7 tosses), and find the differences between the way of programming.
We actually wanted to add another connection with theory, namely, to verify if the
formula
\begin{equation}
V(X)=E((X-E(X))^2)=E(X^2)-E(X)^2
\label{variance}
\end{equation}
can be reflected in the two solutions.

As another activity, a third program was given (see Fig.~\ref{6c.py}) to prepare the last
topic, namely, some steps towards the central limit theorem (CLT).
\begin{figure}
\lstinputlisting[language=Python]{../progs/en_US/6c.py}
\caption{A third program in the sixth sheet (translated into English)}
\label{6c.py}
\end{figure}
Here we provide the questions to this third program:
\begin{enumerate}
\item What computation is performed by this program?
\item Consider following modifications:
\begin{lstlisting}[language=Python,firstnumber=2]
Ω = set(itertools.product({1, 2, 3, 4, 5, 6}, repeat=4))
possible_values = range(30)
\end{lstlisting}
\begin{lstlisting}[language=Python,firstnumber=8]
  heads[sum(ω)] += 1
\end{lstlisting}
Explain the output of the modified program.
\item We are about to plot the obtained data. So we change line 10 as follows:
\begin{lstlisting}[language=Python,firstnumber=10]
for w in zip(*(possible_values, probabilities)):
  print(f"{w[0]},{w[1]}")
\end{lstlisting}
Copy the output of this program in a spreadsheet software and create an XY diagram
to get some graphical output.
\item Extend the program with these two additional lines:
\begin{lstlisting}[language=Python,firstnumber=12]
distribution = [sum(probabilities[:h]) for h in possible_values]
print(list(zip(*(possible_values, distribution))))
\end{lstlisting}
Explain the result.
\end{enumerate}
In fact, the variable naming was somewhat incorrect here (because \texttt{heads} 
actually refers to the sums of the outcomes), but we wanted to make very little
changes on the code.

Among the authors there was a quite a long discussion on getting the output in
the best fashion. The Python construction \texttt{list(zip(*(...)))} is a kind
of compromise between beautiful output with much code and simple output with less code,
however, this is something non-trivial in Python.
Finally a graphical output was also expected---students should be familiar with
popular spreadsheet software and visualize data in an easier way than programming.

For the first part, all solutions were correct: three of them mentioned that
for the second code only one loop is required, so it is somewhat more effective.
In fact all of the turned-in solutions lacked of the wow feeling that we used
the formula (\ref{variance}). One student reported
experiments to learn if \texttt{**} is a typo or not---finally the correct interpretation
was found: it is \textit{the power sign}.

For the second part, the students successfully identified the purpose of the third program
(that is, to find all probabilities for the different number of heads for 5 tosses), and of the changed one
(to find all probabilities of the sums of the outcomes for 4 dice). Both the density
function and the cumulative distribution were obtained correctly, and for the former,
a graphical output was always attached. (One student created a graph for the latter as well.)
We also learned that, in the case of three students, the density function of the
\textit{discrete} binomial distribution
was incorrectly identified, namely, as a \textit{continuous} normal distribution.

We also noted that the modification \texttt{possible\_values = range(30)} aroused interest
among the students,
namely, to find the meaning of the number 30. (In fact, only the sums $4,5,\ldots,24$
had a meaning, but for simplicity we wanted to avoid any manipulation of the indexes of
the arrays.)

\subsection*{Sheet 7}

In the final sheet we wanted to point towards CLT, and thus we gave two programs:
they can be seen in Fig.~\ref{7a.py} and \ref{7b.py}).

\begin{figure}
\lstinputlisting[language=Python]{../progs/en_US/7a.py}
\caption{First program in the seventh sheet (translated into English)}
\label{7a.py}
\end{figure}

\begin{figure}
\lstinputlisting[language=Python]{../progs/en_US/7b.py}
\caption{Second program in the seventh sheet (translated into English)}
\label{7b.py}
\end{figure}

In fact, these two programs are almost the same, just the outputs differ: the second
one results in a nice diagram via \texttt{matplotlib} (see Fig.~\ref{72a}). We provided an alternative
way to plot a vertical bar chart with the following suggestion:
\begin{lstlisting}[language=Python,firstnumber=10]
  p.bar(range(1, max_tosses * 6 + 1), probabilities[1:], alpha=0.5)
\end{lstlisting}
This is mathematically more correct because the discrete values are highlighted
(see Fig.~\ref{72b}).
On the other hand, to point towards CLT, also a continuous diagram makes sense.

\begin{figure}
\centering
\subfloat[An XY diagram via \texttt{pyplot.plot}]{%
\resizebox*{6cm}{!}{\includegraphics{72a}}\label{72a}}\hspace{5pt}
\subfloat[A bar chart via \texttt{pyplot.bar}]{%
\resizebox*{6cm}{!}{\includegraphics{72b}}}
\caption{Graphical outputs of the second program in the seventh sheet} \label{sample-figure}
\end{figure}

Technically, we asked the students to use
Sage worksheets at \url{cocalc.com} since it provides a simple interface to obtain
the graphical result.

In this closing activity we requested to obtain as many mathematical properties of the bar chart
as possible. We asked the students to generalize their experiences. We wanted them to
focus on using notions which are also familiar in a secondary school context. We clarified
that notions like sequence, limit, function, density function, expected value, variance,
standard deviation and normal distribution are all allowed.
